---
title: "The Temporal Dynamics of Emotional Energy"
author: "Matteo Tiratelli"
date: "19/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

For many of the earliest sociologists, the emotional dynamics provoked by crowds and group interaction were the heart of the new discipline. Today, the most famous of these theories is Durkheim’s foundational work on *effervescence collective* – a force or energy generated by ritual interactions, which helps to unify the group and becomes invested in particular ‘sacred’ objects. These ideas have left a lasting impression on the modern sociology of emotions and, in particular, on the enormously influential work of Randall Collins (especially Collins 2004a, 2008). In Collins’s (2012) recent research on violence, he not only makes emotional energy (visceral feelings of collective identity and shared focus of attention) a central concept, but also argues that the temporal profile of emotional energy is characterised by a particular shape: explosion, plateau and slow dissipation (reproduced in Figure 1 below). His primary evidence for this is an ingenious attempt to measure the solidarity produced by conflict by counting public displays of U.S. flags in the aftermath of the September 11th attacks (Collins 2004b). This is a relatively passive form of group identification in the sense that, once the flag is up, you have to make an effort to take it down. But Collins is clear that he suspects that other forms of emotional energy will also follow this pattern of explosion, plateau and dissipation (Collins 2012: 14).

```{r , echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(cowplot)
library(magick)
library(gtrendsR)

## Collect data

search <- tibble(Event = c('Charlie Hebdo shooting','November 2015 Paris attacks',
                           '2016 Nice truck attack','2016 Berlin truck attack',
                           'Boston Marathon bombings','Manchester Arena bombing',
                           '2016 Brussels bombings', '2017 Barcelona attacks'),
                 TopicSearchCode = c('/m/012m7z4x','/g/11bwpyd6bz','/g/11c0rkxllm','/g/11c2nmxm9z',
                                     '/m/0t4zkny','/g/11df0whly0','/g/11cm88pk6n','/g/11f03_cv65'), # These were extracted manually from Google Trends
                 Country = c('FR','FR','FR','DE','US','GB','BE','ES'),
                 Date = c("2015-01-07 2015-09-07", "2015-11-13 2016-07-13", "2016-07-14 2017-03-14", "2016-12-19 2017-08-19",
                          "2013-04-15 2013-12-15", "2017-05-22 2018-01-22","2016-03-22 2016-11-22","2017-08-17 2018-04-17")) # Date of attack plus 8 months

# Download gtrends data to list

results <- vector("list", length = length(search$Event))

for (i in 1:length(search$Event)){
  gtrends(keyword = search$TopicSearchCode[i],
          geo = search$Country[i],
          time = search$Date[i]) -> results[[i]]
}

# Compile results into single tibble

Output <- data.frame(date = as.Date(as.character('2012-01-01')),
                     hits = as.character(2010),
                     keyword = as.character('2010'),
                     geo = as.character('2010'),
                     row.number = as.integer('2010'),
                     stringsAsFactors=FALSE)

for (i in 1:length(search$Event)){
  results[[i]]$interest_over_time[1:4] -> temp
  temp %>%
    mutate(row.number = row_number()) -> temp
  bind_rows(Output, temp) -> Output
}

Output <- as_tibble(Output[-1,])
Output <- merge(Output, search[,1:2], all.x = TRUE, by.x = 'keyword', by.y = 'TopicSearchCode')
Output$Event <- as.factor(Output$Event)


## Reproduce Collins graph

temp = tempfile(fileext = ".png")
download.file("https://raw.githubusercontent.com/MatteoTiratelli/TimeDynamicsEmotionalEnergy/main/CollinsCurve.png", destfile = temp, mode = 'wb')

Collins <- ggplot(Output, aes(x=row.number, y=as.numeric(hits))) +
  theme_classic() + 
  xlab(NULL) +
  ylab(NULL) +
  scale_x_continuous(breaks = c(14,90,180), 
                     labels = c('2 weeks', '3 months', '6 months')) +
  labs(title = "Figure 1: The Collins model of emotional energy over time",caption = "Notes: Public display of U.S. flags after the September 11th attacks. Graph reproduced from Collins (2012), figure 13. Data from \nCollins (2004).") +
  theme(plot.title.position = "plot",
    plot.caption = element_text(hjust = 0),
        plot.caption.position =  "plot",
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))

ggdraw() +
  draw_image(temp, x = -0.01, y = 0.03, scale = 0.8) +
  draw_plot(Collins)

```

Unfortunately, there has been very little work following up on these ideas and attempting to provide other measures of emotional energy over time. The supposition that passive forms of identity display are a good proxy for emotional energy in general, or for conflict solidarity in particular, therefore remains unproven. In order to test this supposition, I compare Collins’s results to the temporal pattern of a more form of active behaviour: online searches, which can be thought of as a proxy for attention towards and emotional investment in a topic. The clearly only addresses one half of what Collins means by 'emotional energy' (shared focus of attention, rather than visceral feelings of collective identity). But attention plays a key role in his theories and he often refers to a sense of 'where the *action* is' as a vital indicator of the presence of emotional energy (Collins 2001).

In order to remain close to Collins’s original example and so to maximise the chances of a confirmatory finding, I use Google Trends data to measure search activity in the aftermath of the eight most deadly recent terror attacks in the United States and western Europe. In each case, the data is a standardised measure of search interest in the particular event in the country which was directly affected by the attack. Rather than following Collins’s explosion-plateau-dissipation model, these examples reveal a very different pattern: dramatic spikes followed by rapid declines. This is true whether we measure searches by the day (Figure 2), or by the hour (Figure 3). Interestingly, this explosive pattern is also much closer to the erratic and jagged patterns we see in counts of protests, riots and strikes over time (Biggs 2005).

```{r , echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE)

## Create new graph

Output$Event <- as.factor(Output$Event)
Output$Event <- fct_relevel(Output$Event, "Charlie Hebdo shooting", 'Boston Marathon bombings', after = Inf)

ggplot(Output, aes(x=row.number, y=as.numeric(hits))) +
  geom_line() + 
  theme_classic() + 
  xlab(NULL) +
  ylab(NULL) +
  scale_x_continuous(breaks = c(14,90,180), 
                     labels = c('2 weeks', '3 months', '6 months')) +
  labs(title = "Figure 2: Emotional attention after terrorist attacks (daily searches)", caption = "Notes: Google Trends Index of searches for each of the events listed, in the country where the event took place. Data extracted \nusing the gtrendsR package and the relevant Google Topics (bundles of related search terms).") +
  facet_wrap(vars(Event), nrow = 4, ncol = 3) +
  theme(plot.title.position = "plot",
        plot.caption.position =  "plot",
        plot.caption = element_text(hjust = 0),        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA),
        legend.background= element_rect(fill = "transparent", colour = NA))

```

```{r , echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(cowplot)

## Compile 'HourlySearches.py' csv output

setwd("/Users/matteo/Downloads/export")

data <- read_csv("0.csv")
names(data)[2] <- 'Terror'
data %>%
  mutate(row.number = row_number()) -> data

for (i in (c(1,2,3,5,6,7))){ # Hourly Google data not available for the Boston Marathon Bombing
  filename = paste0(i,".csv")
  read_csv(filename) -> temp
  names(temp)[2] <- 'Terror'
  temp %>%
    mutate(row.number = row_number()) -> temp
  bind_rows(data, temp) -> data
}

data$Event <- as.factor(data$Event)
data$Event <- fct_relevel(data$Event, "Charlie Hebdo shooting", after = Inf) # Move Charlie Hebdo to the bottom of the facet grid to emphasise its unique pattern

## Plot and save faceted graph

ggplot(data, aes(x=row.number, y=as.numeric(Terror))) +
  geom_line() + 
  theme_classic() + 
  xlab(NULL) +
  ylab(NULL) +
  labs(title = "Figure 3: Emotional attention after terrorist attacks (hourly searches)", caption = "Notes: Hourly searches for 'terrorism' (translated into the relevant language) in the country directly affected by the attack. Data \nextracted using the pytrends package.") +
  facet_wrap(vars(Event), nrow = 4, ncol = 3) +
  theme(plot.title.position = "plot",
        plot.caption.position =  "plot",
        plot.caption = element_text(hjust = 0),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA),
        legend.background= element_rect(fill = "transparent", colour = NA))
```

There are several ways of interpreting the stark differences between these two patterns (explosion-plateau-dissipation vs. spike-decline). At the level of theory, one could infer that the two elements of emotional energy (attention and feelings of collective identity) simply have different temporal dynamics. Empirically, the difference between these results might suggest that the September 11 attacks were relatively unusual in their ability to generate a sustained state of heightened emotional energy. Or, at the level of methodology, it could be that the two proxies (public displays of identity and private search activity) are simply measuring different things entirely.

Each of these interpretations points to the need for further research which tries to bridge the gap between theories of emotion and measurement. This is a daunting undertaking. All I can do here is suggest some initial steps. First, we need clearer, operationalisable definitions of emotional energy (a concept which has come in for its share of criticism over the years, e.g. Felson 2009). This will require distinguishing it from emotions more broadly and may include stronger differentiation between its two component parts. Second, we need more theoretical work on the link between active displays of emotional energy and its passive or latent forms. We may be able to find clues in Durkheim's work the embodiment of *effervescence collective* in objects, symbols and relationships, and, of course, in Collins's idea of 'interation ritual chains' (Collins 2004a). Third, the measurement of emotions is clearly in its infancy and it may be that we have to rely on proxy measures for a long time yet. But we nevertheless should try to compare our findings of the temporal dynamics of emotional energy across a population, with other attempts to measure the time profile of emotions at the individual level (how long does grief last?). Various sources suggest themselves for the former: longitudinal data (if it is collected regularly enough), repeated cross-sectional data (if appropriate survey questions can be found), and the use of Twitter/Facebook profile picture frames (a phenomenon which closely matches Collins's original data). The latter is more properly the domain of psychologists and thus represents a fertile ground for interdisciplinary collaboration.

As Collins's work on violence indicates, the study of emotions has huge potential to illuminate many other areas of social life. The next phase is to begin the rigorous quantitative investigation of those ideas, not as a replacement for theoretical and qualitative work but, hopefully, as a valuable supplement.


All data and code are available from [https://github.com/MatteoTiratelli/TimeDynamicsEmotionalEnergy](https://github.com/MatteoTiratelli/TimeDynamicsEmotionalEnergy).

### Bibliography

* Biggs, Michael. ‘Strikes as Forest Fires: Chicago and Paris in the Late Nineteenth Century’. American Journal of Sociology 110, no. 6 (2005): 1684–1714.
* Collins, Randall. 'Social Movements and the Focus ofEmotional Attention' in Passionate Politics. University of Chicago Press, 2001.
* Collins, Randall. Interaction Ritual Chains. Princeton University Press, 2004a.
* Collins, Randall.‘Rituals of Solidarity and Security in the Wake of Terrorist Attack’. Sociological Theory 22, no. 1 (2004b): 53–87.
* Collins, Randall. Violence: A Micro-Sociological Theory. Princeton University Press, 2008. 
* Collins, Randall. ‘C-Escalation and D-Escalation: A Theory of the Time-Dynamics of Conflict’. American Sociological Review 77, no. 1 (2012): 1–20.
* Felson, Richard 'Is violence natural, unnatural, or rational?'. The British Journal of Sociology 60, no. 3 (2009): 577-585.